\Chapter{Introduction}{intro}

\Section{Scope}{scope}

This document is a specification for the Slang programming language.

Slang is language primarily designed for use in \SpecDef{shader programming}: performance oriented GPU programming for real-time graphics.

\Section{Conformance}{conformance}

This document aspires to specify the Slang language, and the behaviors expected of conforming implementations, with a high level of rigor.

The open-source Slang compiler \emph{implementation} may deviate from the language as specified here, in a few key ways:

\begin{itemize}
    \item{The implementation is necessarily imperfect, and can have bugs}
    \item{The implementation may not fully support constructs specified here, or their capabilities may not be as complete as what is required by the specification}
    \item{The implementation may support certain constructs that are experimental, deprecated, or are otherwise intentionally unspecified (or even undocumented)}
\end{itemize}

Where possible, this document will call out known deviations between the language as specified, and the current implementation in the open-source Slang compiler.

\Section{Document Conventions}{conventions}

The key words \textbf{must}, \textbf{must not}, \textbf{required}, \textbf{shall}, \textbf{shall not}, \textbf{should}, \textbf{should not}, \textbf{recommended}, \textbf{may}, and \textbf{optional} in this document are to be interpreted as described in \href{https://www.ietf.org/rfc/rfc2119.txt}{RFC 2119}.

The word "descriptive" in this document is to be interpreted as synonymous with "non-normative."

\SubSection{Typographical Conventions}{typographical}

This document makes use of the following conventions to improve readability:

\begin{itemize}
    \item{When a term is introduced and defined, it will be in italics. For example: a \SpecDef{duck} is something that looks like a duck and quacks like a duck.}
    \item{Code fragments in Slang or other programming languages use a fixed-width font. For example: \code{a + b}.}
    \item{References to language keywords are in bold [TODO: fix kw typeface description]. For example: \kw{if}}
\end{itemize}

\SubSection{Callouts}{callouts}

This document uses a few kinds of \SpecDef{callouts}, which start bold word indicating the kind of callout. For example, the following is a note:

\begin{Note}
Notes are rendered like this.
\end{Note}

The kinds of callouts used in this document are:

\begin{itemize}
    \item{\textbf{Example}s provide non-normative examples of how to use language constructs, or of their semantics.}
    \item{\textbf{Incomplete} callouts are non-normative markers of places where this document is lacking important information, or may be inaccurate.}
    \item{\textbf{Legacy} callouts are described below.}
    \item{\textbf{Limitation} callouts are used to provide non-normative discussion of how the Slang compiler implementation may deviate from the language as specified here.}
    \item{\textbf{Note}s give non-normative information that may help readers to understand and apply the normative information in the document.}
    \item{\textbf{Rationale} callouts provide non-normative explanations about *why* certain constructs, constraints, or rules are part of the language.}
    \item{\textbf{Syntax} callouts are described below.}
    \item{\textbf{Warning}s give non-normative information about language rules or programming practices that may lead to surprising or unwanted behavior.}
\end{itemize}

\SubSection{Legacy Features}{legacy}

Some features of the Slang language are considered \SpecDef{legacy} features.
The language supports these constructs, syntax, etc. in order to facilitate compatibility with existing code in other GPU languages, such as HLSL.

\textbf{Legacy} callouts are used for the description of legacy features. These callouts are normative.

\SubSection{Grammar Productions}{grammar}

The syntax and lexical structure of the language is described \textbf{Syntax} callouts using a notation similar to EBNF.

\begin{itemize}
    \item{\SynRef{Nonterminals} have names in camel case, and are rendered in italics. E.g., \SynRef{CallExpression}.}
    \item{\SynRef{Terminals} in the syntax, both punctiation and keywords are rendered in a fixed-width font. E.g., \code{func}.}
    \item{A question mark as a suffix indicates that the given element is optional. E.g., \SynRef{Expression}?}
    \item{An asterisk as a suffix indicates that a given element may be repeated zero or more times. E.g., \SynRef{Modifier} *}
    \item{A plus as a suffix indicates that a given element may be repeated one or more times. E.g., \SynRef{AccessorDeclaration}+}
    \item{Parentheses in plain text (not fixed-width) are used for grouping. E.g., \SynRef{Expression} (`,` \SynRef{Expression})}
\end{itemize}

\begin{Incomplete}
Also need conventions for how code points are named in productions in the lexical structure.    
\end{Incomplete}


\Section{The Slang Standard Library}{stdlib}

Many constructs that appear to users of Slang as built-in syntax are instead defined as part of the \SpecDef{standard library} for Slang.
This document may, of neccessity, refer to constructs from the Slang standard library (e.g., the \code{Texture2D} type) that are not defined herein.

This document does not provide a normative definition of all of the types, functions, attributes, modifiers, capabilities, etc. that are defined in the Slang standard library.
The normative reference for standard-library constructs *not* defined in this document is the Slang Standard Library Reference.

\begin{Incomplete}
Need to turn that into a link.
\end{Incomplete}

In cases where a language construct is described in \emph{both} this specification and the Standard Library Reference, the construct has all the capabilities and restrictions defined in either source.
In cases where the two sources disagree, there is a correctness issue in one or the other.

\Section{Normative References}{references}

\begin{Incomplete}
This section should list any external documents/specifications/etc. that are necessary in order to correctly interpret the information in this specification.
\end{Incomplete}